Tropical wet secondary forest regeneration patterns

### 2022-11-09 article/methods.Rmd
bioRxiv,poster

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. `r RgoogleMaps::getGeoCode("cabo matapalo")`) near the Piro Biological Station run by Osa Conservation.
(See *@taylor15* for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the vulnerably dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation, © 2009 Google. Map GIS data and photo credit: Max Villalobos, Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```

@@

Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer *(Forestry Suppliers, Inc.)*, as an average of four readings taken facing each cardinal direction.

<!-- ### [Topography] -->

The slope of the forest floor was measured using a rangefinder *(Bushnell, Forestry Suppliers, Inc)* to measure the distance the diagonal between two plot corners and triangulate the observation angle.

@@

Aboveground biomass (AGB) was calculated at the individual stem level using the allometric equation by *@chave15* and later centered at the plot level.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: 
$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
<!-- Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org). -->
<!-- *[References used for this step are given in *Table S1*.] * -->
Values were updated using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations. 
<!-- However, these updates did not appear to change the main results of this study. -->

>>>>

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. `r RgoogleMaps::getGeoCode("cabo matapalo")`) near the Piro Biological Station run by Osa Conservation.
(See *@taylor15* for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the vulnerably dry-adapted species from the Guanacaste region
<!-- @perezcordero02; -->
*[@hulshof20]* grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

```{r map, fig.width=5, fig.height=5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation, © 2009 Google. Map GIS data and photo credit: Max Villalobos, Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```

@@

Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer *(Forestry Suppliers, Inc.)*, as an average of four readings taken facing each cardinal direction.
<!-- ### [Topography] -->
The slope of the forest floor was measured using a rangefinder *(Bushnell, Forestry Suppliers, Inc)* to measure the distance the diagonal between two plot corners and triangulate the observation angle.

@@

Aboveground biomass (AGB) was calculated at the individual stem level using the allometric equation by *@chave15* and later centered at the plot level.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12]*.
<!-- [@khan20]* -->
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: 
$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
<!-- Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org). -->
<!-- *[References used for this step are given in *Table S1*.] * -->
Values were updated using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations. 
<!-- However, these updates did not appear to change the main results of this study. -->

### 2022-11-09 article/results.Rmd
bioRxiv,poster

Aboveground biomass did not tend to change with distance to primary forest edge 
(Fig \@ref(fig:biomass)a), despite significant trends with underlying related variables 
(Fig \@ref(fig:biomass)).
Median plot wood density tended to increase marginally significantly 
(`r glue("p=",round(as.numeric(wd$pval),3))`) 
with edge distance by 
`r wd$statPrint[[1]]$estimate[[2]] %>% rstatix::p_format()` per m 
<!-- ± `r wd$statPrint[[1]]$std.error[[2]] |> round(3) #formatting...` -->
explaining 
*`r "~10"`* % of variance 
among median distance strata values
(Fig \@ref(fig:biomass)b),
while tissue carbon did not change significantly with distance to forest edge 
(Fig \@ref(fig:biomass)c).
Plot stem density also tended to increase significantly 
(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
with edge distance but more strongly and non-linearly, by 
~`r stems12$statPrint[[1]]$estimate[[2]]` ± 
`r stems12$statPrint[[1]]$std.error[[2]]` per m 
(Fig \@ref(fig:biomass)d).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
(Fig \@ref(fig:biomass)e).
Notably, tree diversity decreased marginally significantly 
(`r glue("p=", divwd$pval[[1]] |> rstatix::p_format())`)
by 
~ `r divwd$statPrint[[1]]$estimate[[2]] |> rstatix::p_round()` 
<!-- ± `r divwd$statPrint[[1]]$std.error[[2]] |> rstatix::p_round()` -->
per g plot wood density, 
yet only explaining
*`r "~4"`* % of variance
(Fig \@ref(fig:biomass)f).

```{r biomass, fig.width=4, fig.height=5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density. Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig1.R"), echo = F); 
fig1
# knitr::include_graphics(here::here("figs/fig1.png"))
```

@@

Overall diversity changed significantly with distance to primary forest edge 
(Fig \@ref(fig:taxa)).
Diversity significantly 
(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
decreased non-linearly by 
~`r abs(entropy12$statPrint[[1]]$estimate[[2]])` ± 
`r abs(entropy12$statPrint[[1]]$std.error[[2]])` per m 
as distance to primary forest edge increased, 
which explained 
*`r "73"`*% of variance 
among distance strata medians 
(Fig \@ref(fig:taxa)a) -- 
while in contrast taxonomic richness increased slightly 
(`r glue("p=",round(as.numeric(richness$pval),3))`) 
and linearly 
by 
`r richness$statPrint[[1]]$estimate[[2]]` ± 
`r richness$statPrint[[1]]$std.error[[2]]` per m 
and had *`r "16"`*% variance explained 
among distance strata 
(Fig \@ref(fig:taxa)b).
Community composition and beta diversity also changed significantly 
(`r glue("p=",ordStat[[5]][1])`) with 
`r ordStat$R2[1] *100`% 
variance among distance strata explained by distance to primary forest edge, 
and the first two principal components explaining 
`r ordTbl$PC1_prop[1] *100`% and 
`r ordTbl$PC2_prop[1] *100`% totaling 
`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
of variance among calculated plot distances 
(Fig \@ref(fig:taxa)d).
Key abundant taxa *`r measure2e`* and *`r measure2f`* showed different responses -- *`r measure2e`* nearly tended to decrease with distance to primary forest edge 
(Fig \@ref(fig:taxa)e),
while *`r measure2f`* decreased marginally significantly 
(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
and linearly by 
`r ficus$statPrint[[1]]$estimate[[2]]` ± 
`r ficus$statPrint[[1]]$std.error[[2]]` per m 
with *`r "~20"`*% variance among distance strata explained 
(Fig \@ref(fig:taxa)f).

```{r taxa, fig.width=5, fig.height=4, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa. White and colored dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig2.R"), echo = F);
fig2
# knitr::include_graphics(here::here("figs/fig2.png"))
```

@@

@@

```{r traits, fig.width=3, fig.height=4, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode. Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig3.R"), echo = F);
fig3
# knitr::include_graphics(here::here("figs/fig3.png"))
```

>>>>

Aboveground biomass did not tend to change with distance to primary forest edge 
(Fig \@ref(fig:biomass)a), despite significant trends with underlying related variables 
(Fig \@ref(fig:biomass)).
Median plot wood density tended to increase marginally significantly 
(`r glue("p=",round(as.numeric(wd$pval),3))`) 
with edge distance by 
`r wd$statPrint[[1]]$estimate[[2]] %>% rstatix::p_format()` per m 
<!-- ± `r wd$statPrint[[1]]$std.error[[2]] |> round(3) #formatting...` -->
explaining 
*`r "~10"`* % of variance 
among median distance strata values
(Fig \@ref(fig:biomass)b),
while tissue carbon did not change significantly with distance to forest edge 
(Fig \@ref(fig:biomass)c).
Plot stem density also tended to increase significantly 
(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
with edge distance but more strongly and non-linearly, by 
~`r stems12$statPrint[[1]]$estimate[[2]]` ± 
`r stems12$statPrint[[1]]$std.error[[2]]` per m 
(Fig \@ref(fig:biomass)d).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
(Fig \@ref(fig:biomass)e).
Notably, tree diversity decreased marginally significantly 
(`r glue("p=", divwd$pval[[1]] |> rstatix::p_format())`)
by 
~ `r divwd$statPrint[[1]]$estimate[[2]] |> rstatix::p_round()` 
<!-- ± `r divwd$statPrint[[1]]$std.error[[2]] |> rstatix::p_round()` -->
per g plot wood density, 
yet only explaining
*`r "~4"`* % of variance
(Fig \@ref(fig:biomass)f).
Additional variables tested, including maximum height and diameter, tended to stay the same across the edge gradient *(data not shown)*.

```{r biomass, fig.width=5, fig.height=6, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (e) canopy light availability, all across distance to primary forest edge gradient, and (f) diversity against wood density. Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig1.R"), echo = F); 
fig1
# knitr::include_graphics(here::here("figs/fig1.png"))
```

@@

Overall diversity changed significantly with distance to primary forest edge 
(Fig \@ref(fig:taxa)).
Diversity significantly 
(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
decreased non-linearly by 
~`r abs(entropy12$statPrint[[1]]$estimate[[2]])` ± 
`r abs(entropy12$statPrint[[1]]$std.error[[2]]) %>% rstatix::p_format()` per m 
as distance to primary forest edge increased, 
which explained 
*`r "73"`*% of variance 
among distance strata medians 
(Fig \@ref(fig:taxa)a) -- 
while in contrast taxonomic richness increased slightly 
(`r glue("p=",round(as.numeric(richness$pval),3))`) 
and linearly 
by 
`r richness$statPrint[[1]]$estimate[[2]] %>% rstatix::p_format()` ± 
`r richness$statPrint[[1]]$std.error[[2]] %>% rstatix::p_format()` per m 
and had *`r "16"`*% variance explained 
among distance strata 
(Fig \@ref(fig:taxa)b).
Community composition and beta diversity also changed significantly 
(`r glue("p=",ordStat[[5]][1])`) with 
`r ordStat$R2[1] *100`% 
variance among distance strata explained by distance to primary forest edge, 
and the first two principal components explaining 
`r ordTbl$PC1_prop[1] *100`% and 
`r ordTbl$PC2_prop[1] *100`% totaling 
`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
of variance among calculated plot distances 
(Fig \@ref(fig:taxa)d).
Key abundant taxa *`r measure2e`* and *`r measure2f`* showed different responses -- *`r measure2e`* nearly tended to decrease with distance to primary forest edge 
(Fig \@ref(fig:taxa)e),
while *`r measure2f`* decreased marginally significantly 
(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
and linearly by 
`r ficus$statPrint[[1]]$estimate[[2]]` ± 
`r ficus$statPrint[[1]]$std.error[[2]]` per m 
with *`r "~20"`*% variance among distance strata explained 
(Fig \@ref(fig:taxa)f).
Other taxa did not change significantly with distance to intact forest *(data not shown)*.

```{r taxa, fig.width=6, fig.height=5, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa. White and colored dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig2.R"), echo = F);
fig2
# knitr::include_graphics(here::here("figs/fig2.png"))
```

@@

<!-- # ```{r taxaList} -->
<!-- # source(here::here("tblTaxa.R"));  -->
<!-- # ``` -->

@@

```{r traits, fig.width=4, fig.height=5, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode. Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig3.R"), echo = F);
fig3
# knitr::include_graphics(here::here("figs/fig3.png"))
```

### 2022-11-09 submit/cover.Rmd
bioRxiv,poster

---
title: ""
output: html_notebook
header-includes:
  - \usepackage{setspace}\doublespace
---

```{r keys}
knitr::opts_chunk$set(echo = F)
title <- "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
journal <- "Journal of Applied Ecology"
```

@@

Please consider this manuscript entitled 
“`r title`” 
as a Research Paper. 
This manuscript presents and discusses data that

support the use of roto-till and cover crop mixtures to alleviate soil compaction and weed pressure, which are very common issues for all farmers as agriculture expands into urban environments.

This manuscript both fits within and also expands the scope of this journal publication. 
The manuscript fits well with the scope because 
it discusses changes to soil functions caused by various types and intensities of mechanical tillage, and informs balancing trade-offs during agricultural management. 
In addition analyzing tillage affects, the manuscript also incorporates discussion of cover crops as a comparable management strategy targeting similar soil functions, and more specifically presents the use of cover crop mixtures, rather than traditional single-species cover cropping. 
Finally, this manuscript places these soil management considerations in an urban agriculture context, which is increasingly relevant to the majority of the world’s population living in urban areas and promotes adaptation to more local food supply chain systems.

Overall, this manuscript represents a solid and interesting contribution to the soil tillage and modern agricultural literature. 
The analysis and discussion is presented as a full manuscript, though it could be possible to contribute this a Short Communication. 
We are also happy to recommend your associate editor 
Elke Noellemeyer 
to be associated with our submission, based on a listed interest in 
agroecology. 
As potential reviewers, we point to 
Kami Pothukuchi and Zhijiang Lu (Wayne State), 
Viashelav Vashenev (RUDN, Russia), 
Zhonqi (Joshua) Cheng (USI, Brooklyn Coll), and 
Rattan Lal (Ohio State). 
Finally, please feel free to contact us for any additional relevant information you may like.

>>>>

---
title: ""
output: 
  # html_notebook
  pdf_document
header-includes:
  - \usepackage{setspace}\doublespace
---

```{r keys, echo=F}
knitr::opts_chunk$set(echo = F)
title <- "Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation"
journal <- "Journal of Applied Ecology"
comparison <- "Edge effects on components of diversity and above-ground biomass in a tropical rainforest, by Razafindratsima et al 2017"
```

@@

Please consider this manuscript entitled 
“`r title`” 
as a Research Paper. 
This manuscript presents and discusses data from a regenerating forested tropical timber plantation that expands typical notions of edge effects, beyond degraded edges and fragmented forests, to highlight how adjacent intact primary forests can actually facilitate the restoration of tree biodiversity and help prioritize ecosystem management efforts.

This manuscript fits directly within the scope of this journal publication, primarily in that it uses an applied setting to address fundamental questions in tropical forest ecology. 
<!-- We use an applied setting, a regenerated tropical timber plantation, to answer fundamental questions about forest community ecology, namely digging into how local management works as the strongest driver of tropical forest carbon, focusing showing data on forest biomass, tree biodiversity, and functional trait recovery. -->
Specifically, it moves beyond aboveground biomass measurements to also evaluate data on tree community composition, as well as some functional traits like height, wood density, tissue carbon, successional stage, and dispersal mode.
As a result, this study does a very robust job of evaluating multiple dimensions of forest restoration, including with respect to this new beneficial edge effect hypothesis.
This study also does a solid job of addressing motivating the globally relevant question of how local management, which is the strongest driver of tropical forest carbon, affects the early stages of recovering forest tree diversity, which can lag many decades behind functional recovery of biomass and taxonomic richness.
Overall, this study appears to closely resemble a recent publication in your journal, `r comparison` under Associate Editor *Natalie Butt*.

The analysis and discussion is presented as a full manuscript, with solid potential to generate discussion with and influence practitioners' perspectives. 
We are also happy to recommend some of your associate editors to be associated with our submission -- 
*P.C Abhilash* based on listed interests in restoring degraded habitats,
*Natalie Butt* based on listed interests in neotropical forest diversity and functions affected by humans,
*Cristina Garcia* based on listed interests in seed dispersal,
*David Moreno Mateos* based on listed interests in accelerating ecosystem restoration,
and
*Sharif Ahmed Mukul* based on listed interests in forest management.
As potential reviewers, we point to 
*María Uriarte* (Columbia U),
*Liza Comita* (Yale U), 
*Marcel Caritá Vaz* (U California - Los Ángeles), 
*Jenny Zambrano* (Washington State U),
*Julia Burton* (Michigan Tech U), and 
*Caroline Farrior* (U Texas - Austin). 
Finally, please feel free to contact us for any additional relevant information you may like.
